A geometric approach to linear and quadratic Chabauty

### 2020-01-29 MS.tex
Added log/exp

\subsubsection{From group schemes to formal groups}
We first recall the definition of a formal group.
\begin{defn}
\label{defn:formalgroup}
Let $R$ be a ring, and $n$ be a non-negative integer. Let $x,y,z$ be vectors of $n$ variables. An $n$-dimensional formal group is an element $F = (F_1,\dots,F_n) \in R[[x,y]]^n$ satisfying $F \equiv x + y \mod (x,y)^2$ and $F(F(x,y),z) = F(x,F(y,z))$. If furthermore $F(x,y) = F(y,x)$, this formal group is said to be commutative.
\end{defn}
\begin{exmp}
\label{exmp:gmformalgroup}
Take $n=1$ and $F(x,y) = x + y + xy = (1+x)(1+y)-1$, also known as the multiplicative formal group. This satisfies associativity as $F(F(x,y),z) = (1+x)(1+y)(1+z)-1 = F(x,F(y,z))$.
\end{exmp}
%For such a formal group, by Hensels lemma it turns out that there is a unique $i_F \in R[[x]]^n$ with $F(x,i_F(x)) = F(i_F(x),x) = 0$ and then by the associativity we find $F(x,0) = F(0,x)$. Such a formal group gives rise to a coproduct $\mu: R[[x]] \to R[[x,y]]$, by sending $x_j to F_j(x,y)$ and extending continously for the $(x)$-adic topology; this works exactly because the constant term of the $F_j$ are all $0$. This coproduct, together with the antipode given by extending $x_j \mapsto i_{F,j}$ continuously and the counit $R[[x]] \to R$ sending $x$ to 0, turns $R[[x]]$ into a group object in the category consisting of $(A,I)$ where $A$ is an $R$-algebra, $I$ is an ideal of $A$ and $A$ is $I$-adically complete as in the following way.

@@

If furthermore $G$ is a group scheme over $\Z_p$, then this formal scheme promotes to a formal group scheme, i.e. for every finite $\Z_p$-algebra $A$ the set $\Hom_{\text{cont}}(\Hat{O}_{G,e},A)$ gets a group structure, functorially in $A$. By functoriality, this is the same [TODO: why exactly?] as an continuous coproduct $\mu: \Hat{O}_{G,e} \to \left(\Hat{O}_{G,e}\right)^{\tensor 2}$. After choosing an isomorphism between $\Hat{O}_{G,e}$ and $\Z_p[[x_1,\dots,x_d]]$, let $F_j \in \Z_p[x,y]$ denote $\mu(x_i)$. Then it is clear that $F = (F_1,\dots,F_d)$ is a $d$-dimensional formal group scheme, and commutative if $G$ was commutative.
\begin{exmp}
\label{exmp:gmtoformal} %TODO: nicely LaTeX Example bla, continued
Take $G = \G_m$ over $\Z_p$, i.e. $\Z_p[u,u^{-1}]$. Then the zero section $e$ is the map sending $u$ to $1$, and we can identify the completion $\Hat{O}_{G,e}$ with the power series ring $\Z_p[[u-1]] \cong \Z_p[[x]]$, sending $u-1$ to $x$. The group structure on $G$ then gives rise to the coproduct $\Z_p[[u-1]] \to \Z_p[[u-1,v-1]], u \mapsto uv$ so the coproduct on $\Z_p[[x]]$ sends $x$ to $uv-1 = (x+1)(y+1) -1 = x + y + xy$. Hence the formal group corresponding to this group scheme is exactly the multiplicative formal group from Example~\ref{exmp:gmformalgroup}.
\end{exmp}

@@

Now we will look at the theory of formal groups, and how that will help us.  over a $\Q$-algebra, Theorem~1 of \citep{honda70} gives the existence of

>>>>

\subsubsection{From group schemes to formal groups}
We first recall the definition of a formal group.
\begin{defn}
\label{defn:formalgroup}
Let $R$ be a ring, and $n$ be a non-negative integer. Let $x,y,z$ be vectors of $n$ variables. An $n$-dimensional formal group is an element $F = (F_1,\dots,F_n) \in R[[x,y]]^n$ satisfying $F \equiv x + y \mod (x,y)^2$ and $F(F(x,y),z) = F(x,F(y,z))$. If furthermore $F(x,y) = F(y,x)$, this formal group is said to be commutative.
\end{defn}
\begin{exmp}
\label{exmp:gmformalgroup}
Take $n=1$ and $F(x,y) = x + y + xy = (1+x)(1+y)-1$, also known as the multiplicative formal group. This satisfies associativity as $F(F(x,y),z) = (1+x)(1+y)(1+z)-1 = F(x,F(y,z))$.
\end{exmp}
\begin{exmp}
\label{exmp:ganformalgroup}
For any $n$, we can take $F(x,y) = x+ y$, also known as the $n$-dimensional additive formal group.
\end{exmp}
%For such a formal group, by Hensels lemma it turns out that there is a unique $i_F \in R[[x]]^n$ with $F(x,i_F(x)) = F(i_F(x),x) = 0$ and then by the associativity we find $F(x,0) = F(0,x)$. Such a formal group gives rise to a coproduct $\mu: R[[x]] \to R[[x,y]]$, by sending $x_j to F_j(x,y)$ and extending continously for the $(x)$-adic topology; this works exactly because the constant term of the $F_j$ are all $0$. This coproduct, together with the antipode given by extending $x_j \mapsto i_{F,j}$ continuously and the counit $R[[x]] \to R$ sending $x$ to 0, turns $R[[x]]$ into a group object in the category consisting of $(A,I)$ where $A$ is an $R$-algebra, $I$ is an ideal of $A$ and $A$ is $I$-adically complete as in the following way.

@@

If furthermore $G$ is a group scheme over $\Z_p$, then this formal scheme promotes to a formal group scheme, i.e. for every finite $\Z_p$-algebra $A$ the set $\Hom_{\text{cont}}(\Hat{O}_{G,e},A)$ gets a group structure, functorially in $A$. By functoriality, this is the same [TODO: why exactly?] as an continuous coproduct $\mu: \Hat{O}_{G,e} \to \left(\Hat{O}_{G,e}\right)^{\tensor 2}$. After choosing an isomorphism between $\Hat{O}_{G,e}$ and $\Z_p[[x_1,\dots,x_d]]$, let $F_j \in \Z_p[x,y]$ denote $\mu(x_i)$. Then it is clear that $F_G = (F_1,\dots,F_d)$ is a $d$-dimensional formal group scheme, and commutative if $G$ was commutative.
\begin{exmp}
\label{exmp:gmtoformal} %TODO: nicely LaTeX Example bla, continued
Take $G = \G_m$ over $\Z_p$, i.e. $\Z_p[u,u^{-1}]$. Then the zero section $e$ is the map sending $u$ to $1$, and we can identify the completion $\Hat{O}_{G,e}$ with the power series ring $\Z_p[[u-1]] \cong \Z_p[[x]]$, sending $u-1$ to $x$. The group structure on $G$ then gives rise to the coproduct $\Z_p[[u-1]] \to \Z_p[[u-1,v-1]], u \mapsto uv$ so the coproduct on $\Z_p[[x]]$ sends $x$ to $uv-1 = (x+1)(y+1) -1 = x + y + xy$. Hence the formal group $F_{\G_m}$ corresponding to this group scheme is exactly the multiplicative formal group from Example~\ref{exmp:gmformalgroup}.
\end{exmp}

@@

Now we will look at the theory of formal groups, and how that will help us. For a $n$-dimensional formal group $F$ over a ring $R$, Proposition~1.1 of \citep{honda70} gives us a canonical $R$-basis $\omega_1,\dots,\omega_n$ of the right invariant differentials; these are elements of $\bigoplus_{j=1}^n R[[x]] \d x_j$. If the formal group is furthermore commutative, then by Proposition~1.3 of \citep{honda70} these are closed, i.e. $d\omega_j = 0$. A $1$-form $d\omega = \sum_{i=1}^n f_i dx_i$ being closed means exactly that $\frac{\partial f_i}{\partial x_j} = \frac{\partial f_j}{\partial x_i}$ for all $i,j$, and then, if $R$ has no torsion, we can formally integrate $\omega$, i.e. write it as $df$ where $f \in (\Q\tensor R)[[x]]$. We make this unique by demanding that $f(0) = 0$. Writing this $f$ as $\sum_{I \in \N^n} a_I x^I$ with $I$ denoting a multi index and $a_I \in \Q\tensor R$, we see that $a_I$, although itself not necessarily lying in $R$, is close: writing $I = (I_1,\dots,I_n)$ we see that for all $j$ we must have $I_j a_I \in R$ as $f_j = \frac{\partial f}{\partial x_j} \in R[[x]]$.

In particular, we write $\log_i$ for the unique element of $(\Q\tensor R)[[x]]$ such that $d\log_i = \omega_i$. Together, these $\log_i$ give an element $\log \in R[[x]]^n$, and by Theorem~1 of \citep{honda70}, this logarithm satisfies $\log(x) = x \mod (x)^2$, and $\log(F(x,y)) = \log(x)+\log(y)$. By the first result, $\log$ has an inverse which we will call $\exp$, also given by powerseries in $(\Q\tensor R)[[x]])$, and also satisfying $\exp(x) = x \mod (x)^2$.
\begin{exmp} %TODO: nicely LaTeX Example bla, continued
Taking $F$ the multiplicative group as in Example~\ref{exmp:gmformalgroup}, Proposition~1.1 of \citep{honda70} gives us $\omega = \frac{1}{x+1}dx$; as a power series this is $\sum_{j \geq 0} (-x)^j$. The formal anti-derivative of this is $\log = \sum_{j \geq 1} \frac{-(-x)^j}{j}$, and Theorem $1$ will tell us what we already know in the context of analysis $\log(x + y + xy) = \log(x)+\log(y)$. Then analysis will also explicitly tell us 
\end{exmp}
